---
title: "STATS 782 Assignment 1"
author: "Chase Robertson - crob873"
date: '15 March 2022'
output: pdf_document
---
\begin{center}
I have read the declaration on the cover sheet and confirm my agreement with it.
\end{center}

# Question 1

## (a)
A sequence of numbers between 1 and 3 using a step of 0.2
```{r}
seq(1, 3, 0.2)
```

## (b)
The sequence of integers between 1 and 4, repeated 3 times.
```{r}
rep(1:4, times=3)
```

## (c)
A vector of strings where each of the first 7 letters is pasted to its index with a space.
```{r}
paste(letters[1:7], 1:7)
```

## (d)
A palindrome of inverse factors of 10, from 0.01 to 1 and back.
```{r}
1 / 10**abs(-2:2)
```

## (e)
A string with all uppercase letters of the alphabet interlaced next to their lowercase.
```{r}
paste0(LETTERS[1:26], letters[1:26], collapse="")
```

# Question 2

## (a)
The probability mass function from 0 to 20 of a Poisson distribution with $\lambda = 5$.
```{r}
plot(dpois(0:20, 5))
```

## (b)
Calculate $P(x=8) + P(x=9) + P(x=10)$ by calculating $P(7<x<=10)$.
```{r}
ppois(10, 5) - ppois(7, 5)
```

## (c)
Calculate $P(x=8) + P(x=9) + P(x=10)$ directly.
```{r}
sum(dpois(8:10, 5))
```

## (d)
Calculate $P(x=8) + P(x=9) + P(x=10)$ by Monte Carlo simulation.
```{r}
# simulate a single independent month of crashes
sim_month <- function() {
  rpois(1, 5)
}

# repeat for 10,000 months
results <- 0
for (i in 1:10000) {
  results[i] = sim_month()
}

# separate months with between 8 and 10 crashes
eight_to_ten <- results[results >= 8 & results <= 10]

length(eight_to_ten) / length(results)
```

# Question 3
A function ```analyse_text``` which takes a single string and returns a named vector containing the number of whitespace-separated words, the length of the longest word, and the mean word length. Punctuation is ignored.
```{r}
analyse_text <- function(x) {
  result <- vector('numeric', 3)
  names(result) <- c('num_words', 'max_word_length', 'mean_word_length')
  
  # remove special characters
  clean_x = gsub('[[:punct:]]', '', x)
  
  # split string into vector of individual words
  words = strsplit(clean_x, " ")[[1]]
  w_i = seq(along=words)

  result['num_words'] = length(words)
  result['max_word_length'] = max(nchar(words[w_i]))
  result['mean_word_length'] = mean(nchar(words[w_i]))
  
  result
}
```

Two tests of the above function.
```{r}
analyse_text("Hello!")
analyse_text("The quick brown fox jumps over the lazy dog.")
```

### EOF